package---
title: "Rat analysis"
output: html_notebook
---


```{r echo=FALSE, results='hide', message=FALSE}
library(RPostgreSQL)
library(knitr)
options(scipen=999,digits=2)
options(dplyr.summarise.inform = FALSE)
knitr::opts_chunk$set(echo = FALSE, cache = TRUE, max.print = -1)
Sys.setenv(PGPASSWORD="xxxx")
con <- dbConnect(RPostgres::Postgres(),
                     dbname="postgres",
                 port="5432",
                 host="xxxx",
                 user="xxxx"
                 )
library(dplyr,warn.conflicts = FALSE)
library(tidyr)
library(tidyverse) 
library(ggplot2)
library(lubridate)
library(readr)
library(ggmap)
library(sf)
library(viridis)
library(tigris)
library(leaflet)
library(mapview)
library(rjson)
```

## Analyze Chicago's rat bating data

Start by importing the rat and and doing counts on the number of request numbers and the number of original request numbers

```{r}
rats_streets_san <- read_csv('~/Desktop/rats/2019_2022.csv')
```

```{r}
#number of service request number
NROW(rats_streets_san$`Service Request Number`)
#number of distinct service request numbers
n_distinct(rats_streets_san$`Service Request Number`)
```


```{r}
rats_streets_san_baited <- rats_streets_san 

```


```{r}
#rename date opened column

rats_streets_san_baited <- rename(rats_streets_san_baited, date=`Date/Time Opened`)

```

```{r}
#add year

rats_streets_san_baited$year <- year(rats_streets_san_baited$date)
```


```{r}
#find totals
total_2019_2022 <- rats_streets_san_baited %>%
  group_by(year) %>% 
  summarize(total = n())

#add category
total_2019_2022$category <- c("total complaints")
```

```{r}
#union baited and totals

graph_data_2019_2022 <- total_2019_2022
```



## 411 rat data

```{r}
#The City of Chicago's 411 agency also released records going back several years related to rat baitings. This will help us get a better understanding of how rat complaints have changed over the decades

rat_data_411 <- read_csv('~/Desktop/rats/csv/2010_2018_rat_complaints.csv')
```

```{r}
nrow(rat_data_411)

n_distinct(rat_data_411$`Service Request No`)
```

```{r}
#standardize date fomrat
rat_data_411$`Created Date` <- dmy(rat_data_411$`Created Date`)
```

```{r}
#create year column
rat_data_411$year <- year(rat_data_411$`Created Date`)
```


```{r}
#filter out original complaints
distinct_rat_complaints <- rat_data_411 %>%
  select(year,`Service Request No`) %>% 
  distinct(year,`Service Request No`) %>% 
  na.omit()
```

```{r}
#tally annual totals
complaints_2010_2018 <- distinct_rat_complaints %>% 
  group_by(year) %>% 
  summarize(total=n())

complaints_2010_2018$category <- c("total complaints")
```

```{r}
#union annual counts with annual bait counts

graph_data_2010_2018 <- complaints_2010_2018
```

```{r}
#union totals from 2018-2018 with 2019-2022

rat_baiting_graph <- union(graph_data_2010_2018,graph_data_2019_2022)
```


```{r}
#create graph of rat complaints
ggplot(data=rat_baiting_graph, aes(x=year, y=total)) + geom_line() + theme_minimal()
```

## Rat map

```{r}
#create new object for map
rats_map <- rats_streets_san_baited
#break up address to prepare for geocoding
rats_map[, 16:17] <- str_split_fixed(rats_map$`Service Request Address`, "<br>", 2)
#isolate geocoding column
rat_complaint_address <- rats_map %>%
  dplyr::select(address=V1,everything())
#join new data field for geocoding
rat_complaint_address$address_concat <-  c(" , Chicago, IL")

#create object to eventually join with geocoded data
rat_map_data <- rat_complaint_address %>%
  mutate(address_full=paste(address, address_concat, sep=""))

#concatenate the fields for geocoding
rat_complaint_address <- rat_complaint_address %>%
  mutate(address_full=paste(address, address_concat, sep="")) %>% 
  dplyr::select(address=address_full) %>%
  distinct(address)
  
```

```{r}
#geocode the rat addresses, all hail the rat king

# remove hashtag to execute — hashtag in place to ensure you don't accidentally start geocoding a batch of coordinates that Google will charge you hundreds of dollars for

# rat_complaints_geocoded <- mutate_geocode(rat_complaint_address,address,output="latlona")
```

```{r}
write_csv(rat_complaints_geocoded,'~/Desktop/rats/rat_addresses_geocoded.csv')
```

```{r}
rat_complaints_geocoded <- read_csv('~/Desktop/rats/rat_addresses_geocoded.csv')
```


```{r}
#join geocoded addresses with rat complaint data

map_data <- rat_map_data %>%
  inner_join(rat_complaints_geocoded,rat_map_data,by=c("address_full"="address...1")) %>% 
  select(-c(address,`Service Request Address`,V2,address_concat,address_full))

#rename geocoded address

map_data <- rename(map_data, address=address...4)

#remove nas for mapping 

map_data <- subset(map_data, !is.na(lat))

#map data for cesar

write_csv(map_data,'~/Desktop/rats/map_data.csv')

```

```{r}
#start building the rat map by importing city shapefiles

chicago_map_shape <- st_read("~/Desktop/rats/shape_files/Boundaries - City/")
chicago_pd_map <- st_read("~/Desktop/rats/shape_files/Boundaries - Police Districts (current)/")
chicago_pd_beat_map <- st_read("~/Desktop/rats/shape_files/police_beat_files/")
chicago_streets_map <- st_read("~/Desktop/rats/shape_files/Street Center Lines/")
chicago_neighborhoods <- st_read("~/Desktop/rats/shape_files/Boundaries - Neighborhoods/")

chicago_pd_beats <- chicago_pd_beat_map["beat_num"]
chicago_pd_boundaries <- chicago_pd_map["dist_num"]
chicago_boundaries <- chicago_map_shape["shape_area"]

chi_neigh <- chicago_neighborhoods["pri_neigh"]
```

```{r}
chi_highways <- chicago_streets_map %>% filter(class == 1 )

chi_arterials <- chicago_streets_map %>% filter(class == 2)

chi_highways <- chi_highways["class"]

chi_arterials <- chi_arterials["class"]

```


```{r}

#project the rat complaint lats and longs

rat_complaints  <- st_as_sf(x = map_data, 
                        coords = c("lon", "lat"),
                        crs = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0") %>% 
  st_transform(crs=st_crs(chicago_pd_beats))

sf::sf_use_s2(FALSE)

#link the lats and longs to the police beats

points_in <- st_join(chicago_pd_beats, rat_complaints, left=T)

```

```{r}
#group by police beat

complaints_by_beat <- points_in %>% 
  group_by(beat_num) %>% 
  summarize(total_complaints=n())
```

```{r}
rat_complaints_by_beat_map <- ggplot(complaints_by_beat) +
  geom_sf(aes(fill=total_complaints),color=NA) +
  coord_sf(datum=NA) +
  labs(title = "Rat complaints by police beat",
       subtitle = "From 2019 to 2022",
       caption = "Source: Chicago Streets and Sanitation",
       fill = "Number of complaints") +
  scale_fill_viridis(option="mako", direction=-1) +
  geom_sf(data = chi_highways,
          inherit.aes = FALSE,
          color = "black",
          size=.8) + 
    geom_sf(data = chi_neigh,
          inherit.aes = FALSE,
          fill=NA,
          size=.2) + 
  theme_void()
```

```{r}
rat_complaints_by_beat_map 
```



#Rat neighborhood maps

```{r}
rat_complaints_neigh  <- st_as_sf(x = map_data, 
                        coords = c("lon", "lat"),
                        crs = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0") %>% 
  st_transform(crs=st_crs(chi_neigh))

sf::sf_use_s2(FALSE)

#link the lats and longs to the police beats

points_in_neigh <- st_join(chi_neigh, rat_complaints_neigh, left=T)
```


### Building code violations for rats

```{r}
rats_building_violations <- read_csv('~/Desktop/rats/rat_violations_buildings.csv')
```

```{r}
#lowercase columns, rename columns, remove NAs from data set

names(rats_building_violations) <- tolower(names(rats_building_violations))

rats_building_violations <- rats_building_violations %>%
  select(everything(),lat=latitude,lon=longitude)

rats_building_violations <- subset(rats_building_violations, !is.na(lat))
```



```{r}
rats_building_violations  <- st_as_sf(x = rats_building_violations, 
                        coords = c("lon", "lat"),
                        crs = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0") %>% 
  st_transform(crs=st_crs(chi_neigh))

sf::sf_use_s2(FALSE)

#link the lats and longs to the police beats

points_in_neigh <- st_join(chi_neigh, rats_building_violations, left=T)



```

```{r}
#drop geometry to prepare for a spreadsheet

rat_building_code_violations <- st_drop_geometry(points_in_neigh)
```


```{r}
#write to CSV

write_csv(rat_building_code_violations, '~/Desktop/rats/building_code_violations_neighborhoods.csv')

```


## Censor phone numbers in 311 rat complaint data set

```{r}
complaints_phone <- read_csv('~/Desktop/rats/neighborhood_rat_complaints.csv')
```

```{r}
#isolate cols
redact_phone <- complaints_phone %>%
  dplyr::select(Neighborhood=pri_neigh, description=`Service Request: Description`, Outcome=Answer, Detail=`Flex Answers Concatenated`, date, address, lon, lat)
```

```{r}
#create regex
phone <- regex("
  \\(?       # optional opening parens
  (\\d{3})   # area code
  \\)?       # optional closing parens
  (?:-|\\ )? # optional dash or space
  (\\d{3})   # another three numbers
  (?:-|\\ )? # optional dash or space
  (\\d{4})   # four more numbers
  ", comments = TRUE)
```

```{r}
#insert the regex to remove the phone numbers from the data set

redact_phone$description <- str_replace_all(redact_phone$description, phone, "XXXXXXXXXX")

```

```{r}
write_csv(redact_phone, '~/Desktop/rats/redacted_phone_numbers.csv')
```


## Suzy B. Wilson ticketing data and geocoding

```{sql connection=con, output.var="suzie_b_rats", echo=FALSE, cache = TRUE}
SELECT * FROM rat_fines
WHERE respondent_business_name ILIKE '%Suzie B%'
OR respondent_business_name ILIKE '%Susie B%'
OR respondent_business_name ILIKE '%Swedlana%'
OR respondent_business_name ILIKE '%Micheal J%'
```

```{r}

#find addresses from old dataset of fines for Suzie B. Wilson and check them again

old_addresses <- suzie_b_rats %>% 
  dplyr::select(violation_address,violation_city,violation_state) %>%
  unique()
  
```


```{r}
#Save object showing Suzie B. Wilson fines from 2010 through 2018

write_csv(suzie_b_rats,'~/Desktop/rats/suzie b wilson/fines_2010_2018.csv')
```


```{r}
#Create object from 686 addresses from tickets issued to Wilson's companies from 2019 through 2022. Gleaned from Excel spreadsheet

wilson_addresses <- read_csv('~/Desktop/rats/suzie b wilson/wilson_addresses.csv')
```


```{r}

### geocode addresses for the tickets issued to Wilson's companies from 2019 through 2022 

#### wilson_geocodes <- mutate_geocode(wilson_addresses,address,output="latlona")

```

```{r}
#clean up column names
wilson_geocodes <- wilson_geocodes %>% 
  select(address=address...1, lon, lat, address_lat_lon=address...4)
```


```{r}
#Save geocoded address to file
write_csv(wilson_geocodes,'~/Desktop/rats/suzie b wilson/address_geocoded.csv')
```


```{r}
#anti-join old addresses with new addresses

unknown_addresses <- old_addresses %>% 
  anti_join(wilson_geocodes, by=c("violation_address" = "address...1")) %>% 
  mutate(violation_city = str_to_title(violation_city)) %>% 
  mutate(address = str_c(violation_address, ', ', violation_city, ', ', violation_state)) %>% 
  select(address)
```



```{r}

## remove hashtags to geocode 148 addresses from 2010-2018 that were not previously geocoded in the 2019-2022 dataset

## unknown_addresses_geocoded <- mutate_geocode(unknown_addresses,address,output="latlona")
```


```{r}
write_csv(unknown_addresses_geocoded,'~/Desktop/rats/suzie b wilson/unknown_addresses_geocoded.csv')
```

```{r}
unknown_addresses_geocoded  <- read_csv('~/Desktop/rats/suzie b wilson/unknown_addresses_geocoded.csv')
```



```{r}
#rename columns
unknown_addresses_geocoded <- unknown_addresses_geocoded %>% 
  dplyr::select(address=address...1,lat,lon,address_lat_lon=address...4)

```


```{r}
#join old geocoded data with new geocoded data
all_wilson_geocodes <- union(unknown_addresses_geocoded,wilson_geocodes)
```


```{r}
#clean up address by removing "Chicago, IL"
all_wilson_geocodes$address  <- str_remove(all_wilson_geocodes$address,", Chicago, IL")
```

```{r}
#join lat lona data with fine data
wilson_fines_2010_2018_geo <- all_wilson_geocodes %>% 
  inner_join(suzie_b_rats, by =c("address" = "violation_address"))

```

```{r}
#remove NAs from geocoded data
wilson_fines_2010_2018_geo <- wilson_fines_2010_2018_geo %>% 
  filter(is.na(lat))
```

```{r}
#save geocoded data to file.
write_csv(wilson_fines_2010_2018_geo,'~/Desktop/rats/suzie b wilson/2010_2018_suzie_data_geo.csv')
```


```{r}
#prep for map. show unique tickets showing total fines and fees for each address from 2010-2018

suzie_b_wilson_map <- wilson_fines_2010_2018_geo %>% 
  select(address,lat,lon,docket_nbr,current_amt_due,total_paid) %>% 
  distinct() %>% 
  group_by(address,lat,lon) %>%
  summarize(total_tickets=n(),total_owed=sum(current_amt_due),total_paid=sum(total_paid)) %>%
  select(address,total_tickets,total_owed,total_paid,lat,lon)

```

```{r}
write_csv(suzie_b_wilson_map, '~/Desktop/rats/suzie_b_wilson_map.csv')
```



```{r}

#using total geocoded data from 2010 to 2018, map each address by neighborhood according to City of Chicago's shapefiles

wilson_map_coded  <- st_as_sf(x = suzie_b_wilson_map, 
                        coords = c("lon", "lat"),
                        crs = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0") %>% 
  st_transform(crs=st_crs(chi_neigh))

sf::sf_use_s2(FALSE)

#link the lats and longs to the neighborhoods

wilson_points_in <- st_join(chi_neigh, wilson_map_coded, left=T)
```

```{r}
#drop the geometry to prepare to summarize the data by neighborhood
wilson_points_in <- st_drop_geometry(wilson_points_in)
```


```{r}

#show the totals per neighborhood for suzie b. wilson tickets from 2010 to 2018

wilson_neighborhood_totals <- wilson_points_in %>% 
  group_by(pri_neigh) %>% 
  summarize(tickets_neigh=sum(total_tickets),total_owed_neigh=sum(total_owed),total_paid_neigh=sum(total_paid)) %>% 
  drop_na(tickets_neigh)

view(wilson_neighborhood_totals)
```


```{r}
#Upload Suzie B. Wilson's fines from 2019_2013

wilson_fines_2019_2023 <- read_csv('~/Desktop/rats/wilson_fines_2019_2023.csv')
```

```{r}
#join these fines with geocodes 
wilson_fines_2019_2023_geo <- wilson_fines_2019_2023 %>%
  inner_join(all_wilson_geocodes,by="address")
```

```{r}
#distinct the fines and group them together
wilson_fines_2019_2023_total <- wilson_fines_2019_2023_geo %>% 
  select(address,lat,lon,docket_nbr,current_amt_due,total_paid) %>% 
  distinct() %>% 
  group_by(address,lat,lon) %>%
  summarize(total_tickets=n(),total_owed=sum(current_amt_due),total_paid=sum(total_paid)) %>%
  select(address,total_tickets,total_owed,total_paid,lat,lon)
  
```


```{r}
#COMBINING DATA SETS
wilson_2019_2023_totals <- wilson_fines_2019_2023_geo %>% 
  select(address,lat,lon,docket_nbr,current_amt_due,total_paid)

wilson_2010_2019_totals <- wilson_fines_2010_2018_geo %>% 
  select(address,lat,lon,docket_nbr,current_amt_due,total_paid)

```

```{r}
#join the two data sets with union
wilson_2010_2023 <- union(wilson_2010_2019_totals,wilson_2019_2023_totals)
```

```{r}
#summarize all data
wilson_totals_2010_2023 <- wilson_2010_2023 %>% 
  select(address,lat,lon,docket_nbr,current_amt_due,total_paid) %>% 
  distinct() %>%
  group_by(address,lat,lon) %>%
  summarize(total_tickets=n(),total_owed=sum(current_amt_due),total_paid=sum(total_paid))
```

```{r}
write_csv(wilson_totals_2010_2023,'~/Desktop/rats/wilson_totals_2010_2023.csv')
```


```{r}
#prepare to assign neighborhood to each address by reprojecting crs
all_wilson_tix <- st_as_sf(x = wilson_totals_2010_2023, 
                        coords = c("lon", "lat"),
                        crs = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0") %>% 
  st_transform(crs=st_crs(chi_neigh))

sf::sf_use_s2(FALSE)

#link the lats and longs to the neighborhoods

wilson_tix_neigh <- st_join(chi_neigh, all_wilson_tix, left=T)
```

```{r}
#add up totals for neighborhood geometry
wilson_geometry_totals_by_neigh <- wilson_tix_neigh %>%
  drop_na() %>% 
  dplyr::select(Neighborhood=pri_neigh, address,total_tickets,total_owed,total_paid) %>% 
  group_by(Neighborhood) %>% 
  summarize(`Total addresses ticketed`=n(),`Total tickets`=sum(total_tickets),`Outstanding fines`=sum(total_owed),`Total paid`=sum(total_paid)) %>% 
  filter(Neighborhood != 'Loop')
```

```{r}
wilson_tix_neigh <- st_drop_geometry(wilson_tix_neigh)
```


```{r}
##save as shapefile
st_write(wilson_geometry_totals_by_neigh, '~/Desktop/rats/wilson_shapefile.shp')
```


## Determine lenghts of time between 311 complaints and response by neighborhood and year

```{sql connection=con, output.var="med_311_response_length_by_year", echo=FALSE, cache = TRUE}
SELECT year, PERCENTILE_CONT(0.5) WITHIN GROUP(ORDER BY difference) AS median
FROM (SELECT sr_number, DATE_PART('year',created_date) AS year, created_date, closed_date, closed_date - created_date AS difference
FROM rat_complaints_311
ORDER BY DATE_PART('year',created_date), closed_date - created_date) AS diff
GROUP BY year

```


```{r}
#shows the median response length by year
med_311_response_length_by_year
```

```{sql connection=con, output.var="response_length_geo", echo=FALSE, cache = TRUE}
SELECT sr_number, street_address, lat, lon, DATE_PART('year',created_date) AS year, created_date, closed_date, closed_date - created_date AS difference
FROM rat_complaints_311
JOIN rat_geocoded ON rat_complaints_311.street_address = rat_geocoded.address
WHERE lat != 'NA'
ORDER BY DATE_PART('year',created_date), closed_date - created_date;

```


```{r}
#assigns neighborhood to each rat complaint
response_length_coded  <- st_as_sf(x = response_length_geo, 
                        coords = c("lon", "lat"),
                        crs = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0") %>% 
  st_transform(crs=st_crs(chi_neigh))

sf::sf_use_s2(FALSE)

#link the lats and longs to the neighborhoods

points_in_response <- st_join(chi_neigh, response_length_coded, left=T)
```

```{r}
#drop geometry
neighborhood_response <- st_drop_geometry(points_in_response)
```

```{r}
#find the difference in the created date and closed date for each complaint 
date_diff <- neighborhood_response %>% 
  mutate(diff_date = (closed_date - created_date))

date_diff <- date_diff %>%
  drop_na()

```

```{r}
#find the median response time for rat comlpaint for each neighborhood and year
by_hood <- date_diff %>%
  group_by(year,pri_neigh) %>%
  summarize(median_diff=median(diff_date))
```


```{r}
#flip the the seconds into days to see the difference by neighborhood
neighborhood_response_rate <- by_hood %>%
  mutate(diff_converted = seconds_to_period(median_diff))

```

```{r}
#save to file
write_csv(neighborhood_response_rate,'~/Desktop/rats/neighborhood_response_rate.csv')
```


### APPENDIX

Chicago's Streets and Sanitation Department supplied the rat complaint data. Chicago's 311 data portal supplied rat complaint data going as far back to 2010, although we only used individual complaint numbers from those years.

Chicago's Finance Department supplied the complete ticketing data after we supplied a list of municipal code violations that the Streets and Sanitation Department supplied to us when we requested for a list of their rat-related tickets. Because one rat-related ticket might contain seven violations, two of which might not be rat-related (someone might get cited for unshoveled snow on the sidewalk in the same ticket they are getting dinged for an overflowing dumpster, for example), we eliminated these non-rat-related municipal codes from the overall list.
